Prison Health

### 2024-02-15 paper/paper.qmd
removed double spacing

---
title: "Racial Force Probe"
author:
  - Monica Sainani
  - Kaavya Kalani
  - Aamishi Avarsekar
thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
date: today
date-format: long
abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations. The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being. Utilizing data collected over the past [insert number of years], this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners. By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being. This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities. Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(ggplot2)
library(readxl)
library(tidyverse)
library(knitr)
library(here)
```

@@

msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016. For our analysis we will not be considering 2001 but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over.

@@

# Results

@@

# Load cleaned data from CSV
health_data_by_year_long_prop <- read.csv("data/analysis_data/cleaned_health_data_by_year.csv", header = TRUE, stringsAsFactors = FALSE)

# Create the plot
ggplot(health_data_by_year_long_prop, aes(x = as.numeric(Year), y = Proportion, color = Disease)) +
  geom_line(size = 1.0) +
  labs(title = "Trends of Diseases by Year",
       x = "Year",
       y = "Proportion of cases",
       color = "Disease") +
  theme_minimal()

@@

# Load cleaned data
ts_data <- read.csv("data/analysis_data/cleaned_covid19_data.csv", header = TRUE, stringsAsFactors = FALSE)

# Create the figure using ggplot2 with solid lines
ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
  geom_line(size = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
  scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases per 100,000 Among
Those in Prison, Prison Staff, and in the General Population: May to September
2020") +
  theme_minimal() +
  geom_vline(xintercept = as.numeric(as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01"))),
             linetype = "dashed", color = "gray") +
  annotate("text", x = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01")),
           y = 150, label = c("April 2020", "May 2020", "June 2020", "July 2020", "August 2020", "September 2020"),
           vjust = -0.5, hjust = 0, color = "gray") +
  scale_color_manual(values = c("blue", "red", "green"))

@@

# Load cleaned data
health_demographic_data_clean <- read_csv("data/analysis_data/cleaned_health_data_gender.csv")

# Plot the proportions graph using geom_bar with position = "dodge"
ggplot(health_demographic_data_clean, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +
  labs(title = "Proportions of Causes of Death by Category",
       x = "Gender",
       y = "Proportion") +
  scale_fill_discrete(labels = cause_mapping) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-race
#| fig-cap: Causes of deaths among the prisoners by race

# Load cleaned data from CSV
race_health_data_long <- read.csv("data/analysis_data/cleaned_health_data_race.csv", header = TRUE, stringsAsFactors = FALSE)

>>>>

---
title: "Prison Health"
author:
  - Monica Sainani
  - Kaavya Kalani
  - Aamishi Avarsekar
thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
date: today
date-format: long
abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations. The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being. Utilizing data collected over the past 15 years, this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners. By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being. This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities. Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(ggplot2)
library(readxl)
library(tidyverse)
library(palmerpenguins)
#### Preamble ####
# Purpose: Read the cleaned dataset to use it for the analysis
# Author: Kaavya Kalani, Monica Sainani, Aamishi Avarsekar
# Date: 14 February 2024
# Contact: kaavya.kalani@mail.utoronto.ca, monica.sainani@mail.utoronto.ca, aamishi.avarsekar@mail.utoronto.ca
# License: MIT
# Pre-requisites: run scripts/01-data_cleaning.R

#### Workspace setup ####
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("here")
# install.packages("ggplot2")
# install.packages("readxl")
# install.packages("gdata")
# install.packages("dplyr")
# install.packages("haven")

library(ggplot2)
library(readxl)
library(tidyverse)
library(knitr)
library(here)
library(gdata)
library(dplyr)
library(haven)
library(knitr)
library(here)
```

@@

msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016. For our analysis we will not be considering 2001, but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over.

@@

# Results
Based on the original paper’s findings, we decided to further investigate the health and general well-being of prisoners and trends of diseases within prison, causes of death and also the trends of COVID-19 across a 7-day period.

@@

# Load cleaned data from CSV

health_data_by_year_long_prop <- read.csv(file=here("data/analysis_data/cleaned_health_data_by_year.csv"), header = TRUE, stringsAsFactors = FALSE)

# Create the plot
ggplot(health_data_by_year_long_prop, aes(x = as.numeric(Year), y = Proportion, color = Disease)) +
  geom_line(linewidth = 1.0) +
  labs(title = "Trends of Diseases by Year",
       x = "Year",
       y = "Proportion of cases",
       color = "Disease") +
  theme_minimal()
```

#Figure 1 
compares shows the cause of death among prisoners due to diseases. In order to accurately investigate, we adjusted the number of deaths caused by a particular disease by the total number of prisoners recorded for that year. This accounts for the differences in the number of deaths recorded between each cause.

By observing the trends in Figure 1, we can see that the deaths due to cancer and heart diseases is significantly higher compared with other causes. The trends for death due to homicide saw a slight decrease between the later years which is opposite to the rate of deaths caused by suicide. While deaths caused by respiratory diseases remain fairly consistent, there is a gradual increase throughout the years. We can also observe that there is a steady decline in the deaths caused by AIDS. Deaths caused by accident remain fairly uncommon with only a very slight increase in 2012. However, the deaths caused by heart disease remained fairly high yet consistent but saw a steep incline after 2012. The deaths caused by cancer has the highest rate. To take a closer look at this, we can investigate the age-group of the prisoners.

# Load cleaned data
ts_data <- read.csv(file=here("data/analysis_data/cleaned_covid19_data.csv"), header = TRUE, stringsAsFactors = FALSE)
# Load cleaned data
ts_data <- read.csv(file=here("data/analysis_data/cleaned_covid19_data.csv"), header = TRUE, stringsAsFactors = FALSE)

#Convert date column to Date type
ts_data$date <- as.Date(ts_data$date, format = "%Y-%m-%d")

# Create the figure using ggplot2 with solid lines
ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
  geom_line(linewidth = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
  scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases") +
  theme_minimal() +
  geom_vline(xintercept = as.numeric(as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01"))),
             linetype = "dashed", color = "gray") +
  annotate("text", x = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01")),
           y = 150, label = c("April 2020", "May 2020", "June 2020", "July 2020", "August 2020", "September 2020"),
          vjust = -0.5, hjust = 0, color = "gray") +
  scale_color_manual(values = c("blue", "red", "green"))

@@

#figure 2

cause_mapping <- c(
  "cancer" = "Cancer",
  "heart_disease" = "Heart Disease",
  "liver_disease" = "Liver Disease",
  "respiratory_disease" = "Respiratory Disease",
  "aids_related" = "Aid related",
  "all_other_illnesses" = "All other illnesses",
  "suicide" = "Suicide",
  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
  "accident" = "Accident",
  "homicide_d" = "Homicide"
)

# Load cleaned data
health_demographic_data_clean <- read_csv(file=here("data/analysis_data/cleaned_health_data_gender.csv"))

# Plot the proportions graph using geom_bar with position = "dodge"
ggplot(health_demographic_data_clean, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +
  labs(title = "Proportions of Causes of Death by Category",
       x = "Gender",
       y = "Proportion") +
  scale_fill_discrete(labels = cause_mapping) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Figure 3 compares the cause of death due to diseases between female and male prisoners. The data in the graph has also been adjusted by proportion of female and male prisoners to account for the low counts of reported diseases in females as well as their population. Both graphs have a similar general unimodal shape. Both sexes have the highest deaths caused by cancer and heart diseases, with males with a slightly higher rate than females. However, the least deaths in females are caused by homicide where its rate is almost negligible. While homicide in males is significantly less compared to other death causes, it is still significantly higher compared to its female counter part. On the other hand, the least deaths in males occurs due to accident. Liver disease and respiratory disease are almost comparable in females whereas for males, there is a steep decline in deaths caused by respiratory diseases in comparison to deaths caused by liver disease.

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-race-2
#| fig-cap: Causes of deaths among the prisoners by race

# Load cleaned data from CSV
race_health_data_long <- read.csv(file=here("data/analysis_data/cleaned_health_data_race.csv"), header = TRUE, stringsAsFactors = FALSE)

Figure 4 compares the causes of deaths in prison with respect to the different races of the prisoners. It can be observed that the cause of deaths are unimodular across all races but they have different peaks in their modes. Amongst Black prisoners, heart diseases have the highest death rate while in Hispanic, White and other races, cancer has the highest death rate. Amongst the identified races, Hispanic prisoners have the highest suicide and homicide rates and also respiratory diseases. It can also be observed that Black prisoners have the highest rate of aids-related deaths, and White prisoners have the highest rate of deaths caused by heart diseases. While deaths caused by drug/alcohol intoxication remains comparatively low across all races, it is the highest for Hispanic prisoners. Amongst other races, it can be observed that suicide is the highest recorded reason for death which could open way to an analyses for racism within prisons as well.

## Discussion

@@
